MS-A1 Formulae and Equations

### 2021-01-21 ms-a1_formulae-and-equations.rmd
Changed example formatting, added MS-A1_b

### Examples {.tabset .tabset-fade .tabset-pills}

#### Example 1 {.tabset .tabset-fade .tabset-pills}

##### Question

@@

##### Show Solution (a)

@@

##### Show Solution (b)

@@

#### Example 2 {.tabset .tabset-fade .tabset-pills}

##### Question

@@

##### Show Solution

@@

>>>>

#### {.tabset .tabset-fade .tabset-pills .unlisted .unnumbered}

##### Hide

##### Example 1 {.tabset .tabset-fade .tabset-pills}

###### Question

@@

###### Solution (a)

@@

###### Solution (b)

@@

##### Example 2 {.tabset .tabset-fade .tabset-pills}

###### Question

@@

###### Solution

@@

## Practice {.tabset .tabset-fade .tabset-pills}

### Questions

If $a=4$, $b=-1$, and $c=-3$, evaluate:

a) $\quad3a$
b) $\quad2a-3b$
c) $\quad2ac$
d) $\quad ac-3c$
e) $\quad\dfrac{a}{2b}$
f) $\quad\dfrac{2ac}{b-c}$
g) $\quad(2c)^2$
h) $\quad2c^2$

### Answers

a) $\quad12$
b) $\quad11$
c) $\quad-24$
d) $\quad-3$
e) $\quad-2$
f) $\quad-12$
g) $\quad36$
h) $\quad18$

# Solving Equations [![](pres.svg){width=16}](https://claytonpollard.github.io/stage6_standard_presentations/MS-A1/b_solving-equations.pdf){target="blank"} {.tabset .tabset-fade .tabset-pills}

* develop and solve linear equations, including those derived from substituting values into a formula, or those developed from a word description

### Equations

::: {.definition}

**Definition 1 - Equation**

A statement asserting that two *expressions* are equal.

For example:

$$\underset{\overset{\uparrow}{\text{expression 1}}}{3n-3}\overset{\underset{\downarrow}{\text{is equal to}}}{=}\underset{\overset{\uparrow}{\text{expression 2}}}{n+1}$$

### Solving Equations

::: {.definition}

**Definition 2**

To **solve** an equation means to find the value of the unknown.

In order to solve an equation we need to rearrange the equation to **isolate** the unknown. We do this by unpacking the equation using **inverse operations**. The inverse operations are performed on *both sides* of the equation to **maintain the balance**.

Once you have found a solution, you can check it is correct by **substituting** back into the original equation.

#### {.tabset .tabset-fade .tabset-pills}

##### Hide

##### Example 1 {.tabset .tabset-fade .tabset-pills}

###### Question

Solve for $x$:

a) $\quad2x-3=5$
b) $\quad8=4x=-2$
c) $\quad\dfrac{x}{4}+7=5$
d) $\quad\dfrac{1}{3}(x+2)=6$

###### Solution (a)

Solve for $x$:

a) $\quad2x-3=5$
b) $\quad8=4x=-2$
c) $\quad\dfrac{x}{4}+7=5$
d) $\quad\dfrac{1}{3}(x+2)=6$

$\begin{aligned}2x-3&=5\\2x-3\color{red}{+3}&=5\color{red}{+3}\\2x&=8\\2x\color{red}{\div2}&=8\color{red}{\div2}\\x&=4\end{aligned}$

###### Solution (b)

Solve for $x$:

a) $\quad2x-3=5$
b) $\quad8=4x=-2$
c) $\quad\dfrac{x}{4}+7=5$
d) $\quad\dfrac{1}{3}(x+2)=6$

$\begin{aligned}8-4x&=-2\\8-4x\color{red}{+4x+2}&=-2\color{red}{+4x+2}\\10&=4x\\10\color{red}{\div4}&=4x\color{red}{\div4}\\\frac{10}{4}&=x\\x&=\frac{10}{4}\\x&=\frac{5}{2}=2.5\end{aligned}$

::: {.note}

It is best practice to have the unknown on the left hand side (LHS) in your answer.

###### Solution (c)

Solve for $x$:

a) $\quad2x-3=5$
b) $\quad8=4x=-2$
c) $\quad\dfrac{x}{4}+7=5$
d) $\quad\dfrac{1}{3}(x+2)=6$

$\begin{aligned}\frac{x}{4}+7&=5\\\frac{x}{4}+7\color{red}{-7}&=5\color{red}{-7}\\\frac{x}{4}&=-2\\\frac{x}{4}\color{red}{\times4}&=-2\color{red}{\times4}\\x&=-8\end{aligned}$

###### Solution (d)

Solve for $x$:

a) $\quad2x-3=5$
b) $\quad8=4x=-2$
c) $\quad\dfrac{x}{4}+7=5$
d) $\quad\dfrac{1}{3}(x+2)=6$

$\begin{aligned}\frac{1}{3}(x+2)&=6\\\frac{1}{3}(x+2)\color{red}{\times3}&=6\color{red}{\times3}\\x+2&=18\\x+2\color{red}{-2}&=18\color{red}{-2}\\x&=16\end{aligned}$

### Repeated Unknowns

If the unknown appears in the equation more than once, we follow these steps:
1. Expand brackets and collect like terms.
2. If the unknown appears on both sides of the equation, remove it from one side using inverse operations.
3. Isolate the unknown and solve the equation.

#### {.tabset .tabset-fade .tabset-pills}

##### Hide

##### Example 2 {.tabset .tabset-fade .tabset-pills}

###### Question

Solve $\quad4(2x+5)-3(x-2)=16\quad$ for $x$.

###### Solution

Solve $\quad4(2x+5)-3(x-2)=16\quad$ for $x$.

$\begin{aligned}4(2x+5)-3(x-2)&=16\\8x+20-3x+6&=16\\5x+26&=16\\5x+26\color{red}{-26}&=16\color{red}{-26}\\5x&=-10\\5x\color{red}{\div5}&=-10\color{red}{\div5}\\x&=-2\end{aligned}$

##### Example 3 {.tabset .tabset-fade .tabset-pills}

###### Question

Solve for $x$:

a) $\quad4x-3=3x+7$
b) $\quad5-3(-1+x)=x$

###### Solution (a)

Solve for $x$:

a) $\quad4x-3=3x+7$
b) $\quad5-3(-1+x)=x$

$\begin{aligned}4x-3&=3x+7\\4x\color{red}{-3x}-3\color{red}{+3}&=3x\color{red}{-3x}+7\color{red}{+3}\\x&=10\end{aligned}$

###### Solution (b)

Solve for $x$:

a) $\quad4x-3=3x+7$
b) $\quad5-3(-1+x)=x$

$\begin{aligned}5-3(-1+x)&=x\\5+3-3x&=x\\8-3x\color{red}{+3x}&=x\color{red}{+3x}\\8&=4x\\8\color{red}{\div4}&=4x\color{red}{\div4}\\2&=x\\x&=2\end{aligned}$

### Fractions

For equations involving fractions, first write all fractions with a **common denominator**.

For example:

* In $\dfrac{2x}{3}=\dfrac{x}{4}$ the lowest common denominator is $\ \ 12$.
* In $\dfrac{3}{x}=\dfrac{5}{2x-1}$ the lowest common denominator is $\ \ x(2x-1)$.

Once the fractions have common denominators you can **equate the numerators**

#### {.tabset .tabset-fade .tabset-pills}

##### Hide

##### Example 4 {.tabset .tabset-fade .tabset-pills}

###### Question

Solve $\quad\dfrac{2-x}{3}=\dfrac{x}{5}\quad$ for $x$.

###### Solution

Solve $\quad\dfrac{2-x}{3}=\dfrac{x}{5}\quad$ for $x$.

$\begin{aligned}\frac{2-x}{3}&=\frac{x}{5}\\\frac{2-x}{3}\color{red}{\times\frac{5}{5}}&=\frac{x}{5}\color{red}{\times\frac{3}{3}}\\\frac{5(2-x)}{15}&=\frac{3x}{15}\\5(2-x)&=3x&\text{(equating the numerators)}\\10-5x&=3x\\10-5x\color{red}{+5x}&=3x\color{red}{+5x}\\10&=8x\\10\color{red}{\div8}&=8x\color{red}{\div8}\\\frac{10}{8}&=x\\x&=\frac{10}{8}\\x&=\frac{5}{4}=1.25\end{aligned}$

##### Example 5 {.tabset .tabset-fade .tabset-pills}

###### Question

Solve $\quad\dfrac{7}{x+1}=\dfrac{3}{x}\quad$ for $x$.

###### Solution

Solve $\quad\dfrac{7}{x+1}=\dfrac{3}{x}\quad$ for $x$.

$\begin{aligned}\frac{7}{x+1}&=\frac{3}{x}\\\frac{7}{x+1}\color{red}{\times\frac{x}{x}}&=\frac{3}{x}\color{red}{\times\frac{x+1}{x+1}}\\\frac{7x}{x(x+1)}&=\frac{3(x+1)}{x(x+1)}\\7x&=3(x+1)&\text{(equating the numerators)}\\7x&=3x+3\\7x\color{red}{-3x}&=3x\color{red}{-3x}+3\\4x&=3\\4x\color{red}{\div4}&=3\color{red}{\div4}\\x&=\frac{3}{4}=0.75\\\end{aligned}$
